---
title: 'Two-Stage PCA Workflow: clusterpca to metapca'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Two-Stage PCA Workflow: clusterpca to metapca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
library(subpca)
```

## Overview

The `subpca` package implements a two-stage hierarchical PCA approach:

1. **Stage 1**: Apply PCA separately to each cluster/block of variables (`clusterpca()`)
2. **Stage 2**: Apply PCA over the combined cluster-wise PCA scores (`metapca()`)

This workflow is particularly useful when your data has natural groupings of variables (e.g., different measurement types, time blocks, or feature categories) and you want to preserve both within-cluster and between-cluster structure.

## Example: Simulated Multi-Block Data

Let's create a dataset with 100 observations and 50 variables, where the variables naturally fall into 5 clusters:

```{r simulate-data}
set.seed(123)
n_obs <- 100
n_vars <- 50
n_clusters <- 5

# Create data with cluster structure
X <- matrix(rnorm(n_obs * n_vars), n_obs, n_vars)

# Add cluster-specific signals
for (i in 1:n_clusters) {
  cluster_vars <- ((i-1)*10 + 1):(i*10)
  # Add a cluster-specific pattern
  X[, cluster_vars] <- X[, cluster_vars] + 
    2 * sin(seq(0, 2*pi, length.out = n_obs)) * rnorm(1)
}

# Define cluster membership
clus <- rep(1:n_clusters, each = 10)
```

## Stage 1: Cluster-wise PCA

First, we apply PCA separately to each cluster of variables:

```{r clusterpca}
# Fit separate PCA to each cluster, keeping 2 components per cluster
cfit <- clusterpca(X, clus = clus, ccomp = 2)

# Examine the structure
print(cfit)

# Number of clusters
length(cfit$fits)

# Components retained per cluster
sapply(cfit$fits, ncomp)
```

## Stage 2: Meta-PCA

Now we perform PCA over the combined cluster-wise scores:

```{r metapca}
# Perform meta-PCA over the cluster PCA scores
mfit <- metapca(cfit$fits, ncomp = 3)

# Examine the meta-PCA fit
print(mfit)

# Total variance explained
cumsum(sdev(mfit)^2) / sum(sdev(mfit)^2)
```

## Projecting New Data

The two-stage model can project new observations through both stages:

```{r project}
# Create hold-out test data with same structure
X_test <- matrix(rnorm(20 * n_vars), 20, n_vars)
for (i in 1:n_clusters) {
  cluster_vars <- ((i-1)*10 + 1):(i*10)
  X_test[, cluster_vars] <- X_test[, cluster_vars] + 
    sin(seq(0, 2*pi, length.out = 20)) * rnorm(1)
}

# Project through the two-stage model
# Note: metapca tracks the original variable indices
test_scores <- project(mfit, X_test)

dim(test_scores)  # 20 observations x 3 meta-components
```

## Understanding the Two-Stage Structure

The key insight is that `metapca` preserves the connection to the original variables:

```{r structure}
# The meta-PCA knows about the original variable groupings
str(mfit$outer_block_indices)

# Each block corresponds to one cluster
names(mfit$outer_block_indices)

# Variables in first cluster
mfit$outer_block_indices[[1]]
```

## Reconstruction

We can reconstruct the original data by reversing through both stages:

```{r reconstruct}
# Reconstruct using all retained meta-components
X_reconstructed <- reconstruct(mfit, comp = seq_len(ncomp(mfit)))

# Compare reconstruction error
reconstruction_error <- norm(X - X_reconstructed, "F") / norm(X, "F")
cat("Relative reconstruction error:", round(reconstruction_error, 4), "\n")
```

## Benefits of the Two-Stage Approach

1. **Interpretability**: Components at each stage have clear meaning
2. **Flexibility**: Different numbers of components per cluster
3. **Efficiency**: Parallel computation of cluster-wise PCA
4. **Robustness**: Local structure preserved within clusters

## Advanced: Using subpca for Convenience

The `subpca()` function wraps this two-stage process:

```{r subpca}
# One-line two-stage PCA
sfit <- subpca(X, clus = clus, ccomp = 2, ncomp = 3)

# Project new data (same result as before)
test_scores2 <- project(sfit, X_test)

# Verify same results
all.equal(test_scores, test_scores2)
```

## Conclusion

The two-stage `clusterpca()` → `metapca()` workflow provides a powerful framework for hierarchical dimensionality reduction that respects the natural structure in your data.
